Deciding Bitvector Termination with Program Synthesis

### 2014-07-01 papers/termination/paper.tex
more on nested loops in sec 5

\begin{definition}[ Nested Loops Termination Formula {\bf [NT]}]
\label{def:deterministic-nonterm-formula}
 \begin{align*}
  \exists R, S . \forall \mathcal{V}, \mathcal{V}' . & G_1(\mathcal{V}) \wedge P_1(\mathcal{V},\mathcal{V}') \rightarrow S(\mathcal{V}',\mathcal{V}') ~ \wedge \\
                                & G_2(\mathcal{V}') \wedge S(\mathcal{V},\mathcal{V}') \wedge T(\mathcal{V}',\mathcal{V}'')\rightarrow S(\mathcal{V},\mathcal{V}'') ~ \wedge \\
				& G_1(\mathcal{V}) \wedge P_1(\mathcal{V},\mathcal{V}') \wedge \neg G_2(\mathcal{V}') \wedge S(\mathcal{V}',\mathcal{V}'') \wedge 
                                P_2(\mathcal{V}'',\mathcal{V}''') \wedge 
                                  R(\mathcal{V}) > 0 \wedge R(\mathcal{V}) > R(\mathcal{V}''') 
 \end{align*} 
\end{definition}

@@

\noindent{\bf Generalised synthesis formula.}
Our design: for each loop always search for lexicographic ranking function + supporting invariant.

@@

Only for certain values of $y$ the loop is non-terminating.
In order to restrict the values that may be assigned to $y$, we replace the non-deterministic choice with a fresh variable $a$ and use formula {\bf [CNT]}
to compute a closed recurrence set $N[a]$ parametrised by $a$. Now $N[a]$ becomes an assertion over $\mathcal{V} \cup \{a\}$.
%, i.e. we under-approximate the transition relation. 
For this example, our technique computes $N[a](x,y,a) = $.%, where $a$ corresponds to the non-deterministic choice.  \\
%The issue with the closed recurrence set is that it yields an incomplete  for loops 
%that have non-deterministic assignments 
%parametric recurrence set

@@

{\bf Sequential loops.}
Discuss {\bf [SNT]}.

{\bf Nested loops.}

>>>>

\begin{definition}[ Nested Loops Termination Formula {\bf [NT]}]
\label{def:nested-term-formula}
 \begin{align*}
  \exists R, S . \forall \mathcal{V}, \mathcal{V}' . & G_1(\mathcal{V}) \wedge P_1(\mathcal{V},\mathcal{V}') \rightarrow S(\mathcal{V}',\mathcal{V}') ~ \wedge \\
                                & G_2(\mathcal{V}') \wedge S(\mathcal{V},\mathcal{V}') \wedge T(\mathcal{V}',\mathcal{V}'')\rightarrow S(\mathcal{V},\mathcal{V}'') ~ \wedge \\
				& G_1(\mathcal{V}) \wedge P_1(\mathcal{V},\mathcal{V}') \wedge \neg G_2(\mathcal{V}') \wedge S(\mathcal{V}',\mathcal{V}'') \wedge 
                                P_2(\mathcal{V}'',\mathcal{V}''') \wedge 
                                  R(\mathcal{V}) > 0 \wedge R(\mathcal{V}) > R(\mathcal{V}''') 
 \end{align*} 
\end{definition}

@@

\noindent{\bf Nested loops.}
For brevity reasons, we illustrate our approach to handling nested loops for only two loops, but the approach can be easily generalised to any number of nested loops:
\begin{lstlisting}[mathescape=true]
while ($G_1$) { 
  $P_1$ 
  while ($G_2$) { $T$ }
  $P_2$
}
\end{lstlisting}

The termination of the outer loop is equivalent to the formula in Definition~\ref{def:nested-term-formula}.
The formula requires an auxiliary assertion $S(\mathcal{V},\mathcal{V}')$ representing a binary relation between the initial states at the inner loop's entry point
and their successors according to the transition relation $T$. If this formula is satisfiable, then $R$ is a ranking function of the outer loop.

In order to find a termination argument for the inner loop, we use {\bf [CT]} where the initial states are given by 
$\neg G_1(\mathcal{V}) \wedge P_1(\mathcal{V},\mathcal{V}') \rightarrow I(\mathcal{V}')$.

@@

Only for certain values of $y$ the loop is non-terminating.
In order to restrict the values that may be assigned to $y$, we replace the non-deterministic choice with a fresh variable $a$ and use formula {\bf [CNT]}
to compute a closed recurrence set $N[a]$ parametrised by $a$. Now $N[a]$ becomes an assertion over $\mathcal{V} \cup \{a\}$.
%, i.e. we under-approximate the transition relation. 
For this example, our technique computes $N[a](x,y,a) = ...$. \\%, where $a$ corresponds to the non-deterministic choice.  \\
%The issue with the closed recurrence set is that it yields an incomplete  for loops 
%that have non-deterministic assignments 
%parametric recurrence set

@@

\noindent{\bf Sequential loops.}
Discuss {\bf [SNT]}.\\

\noindent{\bf Nested loops.}

\subsection{Generalised synthesis formula.}
Our design: for each loop always search for lexicographic ranking function + supporting invariant OR recurrence set.

Note: given a program with multiple loops (sequential and nested), we could potentially compute termination/non-termination 
arguments for all of them in the same formula. Currently, our tool generates a separate formula for each loop.
